The Importance of Initial Conditions in Seasonal Predictions of Antarctic Sea Ice

### 2025-12-04 analysis/paper/access-ice.tex
Renders

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sn-mathphys-num,
  referee,lineno]{sn-jnl}

\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

%%%% Standard Packages

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%

%%%%

\raggedbottom
\usepackage[section]{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The Importance of Initial Conditions in Seasonal Predictions of Antarctic Sea Ice},
  pdfauthor={Elio Campitelli; Ariaan Purich; Julie Arblaster; Eun-Pa Lim; Matthew C. Wheeler; Phillip Reid},
  pdfkeywords={sea ice, seasonal predictability, initial
conditions, forecasting},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[The Importance of Initial Conditions in Seasonal Predictions of
Antarctic Sea Ice]{The Importance of Initial Conditions in Seasonal
Predictions of Antarctic Sea Ice}

% author setup
\author*[1,2]{\fnm{Elio} \sur{Campitelli}}\email{elio.campitelli@monash.edu}\author[1,2]{\fnm{Ariaan} \sur{Purich}}\author[1,2]{\fnm{Julie} \sur{Arblaster}}\author[3]{\fnm{Eun-Pa} \sur{Lim}}\author[3]{\fnm{Matthew C.} \sur{Wheeler}}\author[3]{\fnm{Phillip} \sur{Reid}}
% affil setup
\affil[1]{\orgdiv{School of Earth, Atmosphere and
Environment}, \orgname{Monash University, Kulin Nations, Clayton,
Victoria, Australia.}}
\affil[2]{\orgdiv{SARC Special Research Initiative for Securing
Antarctica's Environmental Future}, \orgname{Clayton, Kulin Nations,
Victoria, Australia}}
\affil[3]{\orgdiv{Research, Bureau of Meteorology, Melbourne,
Australia}}

% abstract

\abstract{Accurate Antarctic sea-ice forecasts are crucial for climate
monitoring and operational planning, yet they remain challenging due to
model biases and complex ice-ocean-atmosphere interactions. The two
versions of the Australian Bureau of Meteorology's ACCESS seasonal
forecast system, ACCESS-S1 and ACCESS-S2, use identical model
configuration and differ only in their initial conditions; primarily in
that ACCESS-S2 does not assimilate sea-ice observations, whereas
ACCESS-S1 does.\\
This provides a convenient opportunistic experiment to assess the role
of initial conditions on Antarctic sea-ice forecasts using more than 20
years of fully coupled simulations with two 9-member ensembles. Our
analysis reveals that both systems experience an extended melt season
and delayed growth phase compared with observations. This leads to a
significant negative sea-ice extent bias, which is corrected only in
ACCESS-S1 by the data assimilation system. The impact of the differing
initial conditions on forecast errors varies dramatically by season:
summer and autumn initial conditions (January-April) provide predictive
skill for up to three months, with February initial conditions being
particularly crucial. In contrast, winter forecasts of the two systems
are statistically indistinguishable after only two weeks. Regional
analysis of forecast skill suggests that this winter predictability
barrier is most dramatic over East Antarctica, where even ACCESS-S1
shows negative skill. These findings highlight the critical importance
of comprehensive year-round sampling in predictability studies and
suggest that operational sea-ice data assimilation efforts should
prioritize the summer-autumn period when initial conditions have maximum
impact on forecast skill.}

% keywords
\keywords{sea ice,  seasonal predictability,  initial
conditions,  forecasting}

\begin{document}
\maketitle

@@

The European Organisation for the Exploitation of Meteorological
Satellites (EUMETSAT) Ocean and Sea Ice Satellite Application Facility
{[}OSI; \citet{OSISAF}{]} based on the SSMIS sensor is another
satellite-derived sea-ice concentration product. It is based on mostly
the same sensors as the NOAA CDR but computed independently using
different algorithms. Figures prepared with this dataset are provided in
the supplementary material and do not differ significantly from the ones
prepared using CDR.

@@

\section{Funding}\label{funding}

This work was supported by ARC SRIEAS Grant SR200100005 Securing
Antarctica's Environmental Future.

\section{Acknowledgments}\label{acknowledgments}

We thank the internal reviewers Bethan White and Xiaobing Zhou for their
comments and feedback.

This work benefited from earlier unpublished work by Laura Davies, Phil
Reid, Andrew G. Marshall.

This research was undertaken with the assistance of resources from the
National Computational Infrastructure (NCI Australia), an NCRIS enabled
capability supported by the Australian Government.

\section{Code and data availability}\label{code-and-data-availability}

The underlying code for this study is available on GitHub:
https://github.com/eliocamp/access-s2\_ice-eval.

Raw data of ACCESS-S1 and ACCESS-S2 hindcast are not available due to
size. Derived datasets required to reproduce the results, including
extent timeseries and error measures, are available in this Zenodo
repository: https://zenodo.org/records/17479538 \citep{campitelli2025}

\section{Competing interests}\label{competing-interests}

All authors declare no financial or non-financial competing interests.

\section{Author contributions}\label{author-contributions}

EC performed the data analysis and wrote the manuscript draft. AP, JA,
EL, MW and PR, performed interpretation of the results, and reviewed and
edited the draft. All authors read and approved the final manuscript.

@@

\appendix  
\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

\section{Supplementary figures}\label{supplementary-figures}

@@

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[bg, manuscript]{copernicus}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\usepackage[inline]{trackchanges}
\usepackage[section]{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The Importance of Initial Conditions in Seasonal Predictions of Antarctic Sea Ice},
  pdfkeywords={sea ice, seasonal predictability, initial
conditions, forecasting},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={red},
  pdfcreator={LaTeX via pandoc}}

\title{The Importance of Initial Conditions in Seasonal Predictions of
Antarctic Sea Ice}

\Author[1, 2]{Elio}{Campitelli}
\Author[1, 2]{Ariaan}{Purich}
\Author[1, 2]{Julie}{Arblaster}
\Author[3]{Eun-Pa}{Lim}
\Author[3]{Matthew C.}{Wheeler}
\Author[3]{Phillip}{Reid}

\affil[1]{School of Earth, Atmosphere and Environment, Monash
University, Kulin Nations, Clayton, Victoria, Australia.}
\affil[2]{ARC Special Research Initiative for Securing Antarctica's
Environmental Future, Clayton, Kulin Nations, Victoria, Australia}
\affil[3]{Research, Bureau of Meteorology, Melbourne, Australia.}

\runningtitle{The Importance of Initial Conditions in Seasonal
Predictions of Antarctic Sea Ice}

\runningauthor{Campitelli et al.}

\correspondence{Elio\ Campitelli\ (elio.campitelli@monash.edu)}

\begin{document}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

%% These dates will be inserted by Copernicus Publications during the typesetting process.

\firstpage{1}
\maketitle
\begin{abstract}
Accurate Antarctic sea-ice forecasts are crucial for climate monitoring
and operational planning, yet they remain challenging due to model
biases and complex ice-ocean-atmosphere interactions. The two versions
of the Australian Bureau of Meteorology's ACCESS seasonal forecast
system, ACCESS-S1 and ACCESS-S2, use identical model configuration and
differ only in their initial conditions; primarily in that ACCESS-S2
does not assimilate sea-ice observations, whereas ACCESS-S1 does.\\
This provides a convenient opportunistic experiment to assess the role
of initial conditions on Antarctic sea-ice forecasts using more than 20
years of fully coupled simulations with two 9-member ensembles. Our
analysis reveals that both systems experience an extended melt season
and delayed growth phase compared with observations. This leads to a
significant negative sea-ice extent bias, which is corrected only in
ACCESS-S1 by the data assimilation system. The impact of the differing
initial conditions on forecast errors varies dramatically by season:
summer and autumn initial conditions (January-April) provide predictive
skill for up to three months, with February initial conditions being
particularly crucial. In contrast, winter forecasts of the two systems
are statistically indistinguishable after only two weeks. Regional
analysis of forecast skill suggests that this winter predictability
barrier is most dramatic over East Antarctica, where even ACCESS-S1
shows negative skill. These findings highlight the critical importance
of comprehensive year-round sampling in predictability studies and
suggest that operational sea-ice data assimilation efforts should
prioritize the summer-autumn period when initial conditions have maximum
impact on forecast skill.
\end{abstract}
\copyrightstatement{The author's copyright for this publication is
transferred to institution/company.}

@@

The European Organisation for the Exploitation of Meteorological
Satellites (EUMETSAT) Ocean and Sea Ice Satellite Application Facility
{[}OSI; \citet{OSISAF}{]} based on the SSMIS sensor is another
satellite-derived sea-ice concentration product. It is based on mostly
the same sensors as the NOAA CDR but computed independently using
different algorithms. Figures prepared with this dataset are provided in
the appendix and do not differ significantly from the ones prepared
using CDR.

@@

@@

\section{Appendix}\label{appendix}

\appendixfigures

@@

\codedataavailability{The underlying code for this study is available on
GitHub: https://github.com/eliocamp/access-s2\_ice-eval. Raw data of +S1
and +S2 hindcast are not available due to size. Derived datasets
required to reproduce the results, including extent timeseries and error
measures, are available in this Zenodo repository:
https://zenodo.org/records/17479538
\citep{campitelli2025}} %% use this section when having data sets and software code available

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontribution{EC performed the data analysis and wrote the
manuscript draft. AP, JA, EL, MW and PR, performed interpretation of the
results, and reviewed and edited the draft. All authors read and
approved the final manuscript.} %% optional section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\competinginterests{The authors declare no competing
interests.} %% this section is mandatory even if you declare that no competing interests are present

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgements}
We thank the internal reviewers Bethan White and Xiaobing Zhou for their
comments and feedback. This work benefited from earlier unpublished work
by Laura Davies, Phil Reid, Andrew G. Marshall. This research was
undertaken with the assistance of resources from the National
Computational Infrastructure (NCI Australia), an NCRIS enabled
capability supported by the Australian Government. This work was
supported by ARC SRIEAS Grant SR200100005 Securing Antarctica's
Environmental Future.
\end{acknowledgements}

\bibliographystyle{copernicus}
